Approximation as Differentiation

### 2025-05-24 main/formal-structures.tex
Expand why LatGal isn't CC.

Vákár and collaborators overcome this by using different interpretations for forward and backward derivatives. We overcome this by using $\CMon$-enriched presheaves; we turn to this in \secref{}, but first we flesh out the first-order setting.

@@

The meet-semilattice homomorphisms from $X$ to $Y$ form a meet-semilattice themselves, with $\meet$ on
homomorphisms defined pointwise, so that $(f \meet g)(x) = f(x) \meet g(x): X \to Y$, and the constant map
$\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$ providing the unit (and both preserving meets). Dually
the join-preserving maps from $Y$ to $X$ have a join-semilattice structure given pointwise by $\join$ and the
constant homomorphism $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to $\bot_X$. Since these two constructions
come in adjoint pairs, $\LatGal$ is enriched in the category $\SemiLat$ of semilattices and semilattice
homomorphisms. The hom-set $\LatGal(X,Y)$ forms a bounded meet semilattice with unit $\top_{X,Y}$ given by the
Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and meet of Galois connections $(f \adj g) \meet (f'
\adj g') = (f \join f') \adj (g \meet g'): X \to Y$.

Another notable property of $\LatGal$ is that the projections $\pi_1: X \times Y \to X$ and $\pi_2: X \times Y
\to Y$, where $X \times Y$ denotes the product of lattices, have both upper and lower adjoints. This means $X
\times Y$, which we shall hereafter write as $X \biprod Y$, acts as both a product and a coproduct, with
projections $\biproj_X$ and $\biproj_Y$ and injections $\biinj_X$ and $\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we will write as $0$, is the
unit for $\biprod$. Such a structure $(\biprod, 0)$ is called a \emph{biproduct} (\secref{biproducts}).
Although $\LatGal$ is not itself Cartesian closed (in particular the set of Galois connections $X \to Y$ does
not in general form a bounded lattice), the biproduct structure will prove crucial for establishing a
Cartesian closed setting for interpreting higher-order programs.

>>>>

Vákár and collaborators overcome this by using different interpretations for forward and backward derivatives. We overcome this by using $\CMon$-enriched presheaves; we turn to this in \secref{higher-order}, but first we flesh out the first-order setting.

@@

The meet-semilattice homomorphisms from $X$ to $Y$ themselves form a meet-semilattice. The operation $\meet$
on homomorphisms is defined pointwise, so that $(f \meet g)(x) = f(x) \meet g(x): X \to Y$, and preserves
meets; the constant map $\top_{X,Y}: X \to Y$ sending any $x \in X$ to $\top_Y$, which also preserves meets,
provides the unit. Dually the join-preserving maps from $Y$ to $X$ have a join-semilattice structure given
pointwise by $\join$ and the constant homomorphism $\bot_{Y,X}: Y \to X$ sending any $y \in Y$ to $\bot_X$.
Since these two constructions come in adjoint pairs, the hom-set $\LatGal(X,Y)$ forms a bounded meet
semilattice with unit $\top_{X,Y}$ given by the Galois connection $\bot_{Y,X} \adj \top_{X,Y}: X \to Y$ and
meet of Galois connections $(f \adj g) \meet (f' \adj g') = (f \join f') \adj (g \meet g'): X \to Y$. Thus
$\LatGal$ is enriched in $\SemiLat$, the category of semilattices and semilattice homomorphisms.

Another notable property of $\LatGal$ is that the projections $\pi_1: X \times Y \to X$ and $\pi_2: X \times Y
\to Y$, where $X \times Y$ denotes the product of lattices, have both upper and lower adjoints. This allows $X
\times Y$, which we shall hereafter write as $X \biprod Y$, to act as both a product and a coproduct, with
projections $\biproj_X$ and $\biproj_Y$ and injections $\biinj_X$ and $\biinj_Y$ given by:

@@

\noindent The trivial 1-point lattice, which is both terminal and initial and we will write as $0$, is the
unit for $\biprod$. Such a structure $(\biprod, 0)$, where $\biprod$ acts as both a product and a coproduct,
is called a \emph{biproduct} (\secref{biproducts}).

$\LatGal$ is not Cartesian closed; the set of Galois connections $X \to Y$ does not in general form a bounded
lattice, for a start. In fact if $\cat{C}$ has biproducts $(\biprod, 0)$ and is Cartesian closed, then every
morphism in $\cat{C}$ is trivial. If $0$ is both initial and terminal, and therefore $0 \times X \iso X$, then
$\Hom{\cat{C}}{X}{Y} \iso \Hom{\cat{C}}{0}{X \Rightarrow Y} \iso 1$ and so any morphism $f: X \to Y$ must be
equal to the composite $X \to 0 \to Y$ of the unique morphisms in and out of $0$. The biproduct structure will
prove crucial for establishing a Cartesian closed setting for interpreting higher-order programs.

### 2025-05-24 main/framework.tex
Start on todo's for new section.
Expand why LatGal isn't CC.

\section{A categorical framework for \GPS}

We will consider a pure, total language, where types can be interpreted as sets. Our goal is to equip the
interpretations of programs with Galois connections relating approximations of their inputs to approximations
of their outputs. Since these lattices of approximations depends on the value being approximated, the
semantics of every type $\tau$ must have both a ``conventional'' component, providing the set $X$ which
interprets $\tau$, plus an associated lattice of approximations $\partial X(x)$ for every point $x \in X$. The
semantics of an expression $e$ of type $\sigma$ with a single free variable of type $\tau$ (with $\tau$ and
$\sigma$ conventionally interpreted by $X$ and $Y$) must then also have two components: a function $f: X \to
Y$ giving the conventional interpretation of $e$, and a family of Galois connections $\partial f(x): \partial
X(x) \to \partial Y(f(x))$ for every $x \in X$ which can be used for slicing over that expression.

@@

@@

\subsection{Interpreting Higher-Order Programs}

\todo{conditions under which $\Fam(\cat{C})$ is Cartesian closed, why we need $\PSh_{\CMon}(\cat{C})$, etc}

>>>>

@@

What we want to do now is understand how to equip the interpretation of a (first-order) program with a Galois
connections relating approximations of its inputs to approximations of its output. Since these lattices of
approximations depend on the value being approximated, the semantics of every type $\tau$ must have both a
``conventional'' component, providing the set $X$ which interprets $\tau$, plus an associated lattice of
approximations $\partial X(x)$ for every point $x \in X$. The semantics of an expression $e$ of type $\sigma$
with a single free variable of type $\tau$ (with $\tau$ and $\sigma$ denoting $X$ and $Y$) must then also have
two components: a function $f: X \to Y$ giving the conventional interpretation of $e$, and a family of Galois
connections $\partial f(x): \partial X(x) \to \partial Y(f(x))$ for every $x \in X$ which can be used for
slicing over that expression.

@@
